Fitness tracking reveals task-specific associations between memory, mental health, and exercise

### 2021-11-05 paper/main.tex
incorporated GMN edits

\begin{abstract}
  Physical exercise can benefit both physical and mental well-being.
  Different forms of exercise (i.e., aerobic versus anaerobic; running
  versus walking, swimming, or yoga; high-intensity interval training
  versus endurance workouts; etc.) impact physical fitness in
  different ways.  For example, running may substantially impact leg
  and heart strength but only moderately impact arm strength. We
  hypothesized that the mental benefits of exercise might be similarly
  differentiated.  We focused specifically on how different forms of
  exercise might relate to different aspects of memory and mental
  health.  To test our hypothesis, we collected (in aggregate) roughly
  a century's worth of fitness data.  We then asked participants to
  fill out surveys asking them to self-report on different aspects of
  their mental health.  We also asked participants to engage in a
  battery of memory tasks that tested their short and long term
  episodic, semantic, and spatial memory performance.  We found that
  participants with similar exercise habits and fitness profiles
  tended to also exhibit similar mental health and task performance
  profiles.  These effects were task-specific in that different
  exercise patterns or fitness characteristics varied with different
  aspects of memory, on different tasks.
\end{abstract}

\section*{Introduction}
Engaging in physical activity (exercise) can improve physical
fitness by increasing muscle strength \citep{RogeEvan93, Lind79,
  CranEtal13, Knut07}, bone density~\citep{ChilEtal12,
  BassRams94, LaynNels99}, cardiovascular
performance~\citep{MaioEtal00, PollEtal00}, lung
capacity~\citep[][although see \cite{RomaEtal16}]{LazoEtal16},
endurance~\citep{WilmKnut03}, and more.  Exercise can also
improve mental health~\citep{Ragl90, MikkEtal17, TaylEtal85,
  DeslEtal09, Call04, PaluSchw00, BassSuzu17} and cognitive performance~\citep{ChanEtal12b,
  BrisEtal02, EtniEtal06, BassSuzu17}.

The physical benefits of exercise can be explained by stress-responses
of the affected body tissues. For example, skeletal muscles that are
taxed during exercise exhibit stress responses~\citep{MortEtal09} that
can in turn affect their growth or atrophy~\citep{SchiEtal13}.  By
comparison, the benefits of exercise on mental health are less direct.  For
example, one hypothesis is that exercise leads to specific
physiological changes, such as increased aminergic synaptic
transmission and endorphin release, which in turn act on
neurotransmitters in the brain~\citep{PaluSchw00}.

Speculatively, if different exercise regimens lead to different
neurophysiological responses, one might be able to map out a spectrum
of signalling and transduction pathways that are impacted by a given
type, duration, and intensity of exercise in each brain region.  For
example, prior work has shown that exercise increases acetylcholine
levels, starting in the vicinity of the exercised
muscles~\citep{ShoeEtal97}.  Acetylcholine is thought to play an
important role in memory formation~\citep[e.g., by modulating specific
synaptic inputs from entorhinal cortex to the hippocampus, albeit in
rodents;][]{PalaEtal21}.  Given the central role that these medial
temporal lobe structures play in memory, changes in acetylcholine
might lead to specific changes in memory formation and retrieval.

In the present study, we hypothesize that (a) different exercise regimens will
have different, quantifiable impacts on cognitive performance and
mental health, and that (b) these impacts will be consistant across
individuals.  To this end, we collected a year of fitness tracking
data from each of 113 participants.  We then asked each participant to
fill out a brief survey in which they self-evaluated and self-reported
several aspects
of their mental health.  Finally, we ran each participant through a
battery of memory tasks, which we used to evaluate their memory
performance along several dimensions.  We searched the data for
potential associations between memory, mental health, and exercise.

@@

\section*{Author contributions}
Concept: J.R.M.  Experiment implementation and data collection: G.M.N.
Analyses: G.M.N., E.C., P.C.F., and J.R.M.  Writing: J.R.M. with input
from all authors.

>>>>

\newpage
\begin{abstract}
  Physical exercise can benefit both physical and mental well-being.
  Different forms of exercise (e.g., aerobic versus anaerobic; running
  versus walking, swimming, or yoga; high-intensity interval training
  versus endurance workouts; etc.) impact physical fitness in
  different ways.  For example, running may substantially impact leg
  and heart strength but only moderately impact arm strength. We
  hypothesized that the mental benefits of exercise might be similarly
  differentiated.  We focused specifically on how different forms of
  exercise might relate to different aspects of memory and mental
  health.  To test our hypothesis, we collected (in aggregate) roughly
  a century's worth of fitness data.  We then asked participants to
  fill out surveys asking them to self-report on different aspects of
  their mental health.  We also asked participants to engage in a
  battery of memory tasks that tested their short and long term
  episodic, semantic, and spatial memory performance.  We found that
  participants with similar exercise habits and fitness profiles
  tended to also exhibit similar mental health and task performance
  profiles.  These effects were task-specific in that different
  exercise patterns or fitness characteristics varied with different
  aspects of memory, on different tasks.  Taken together, these
  findings provide foundational work for designing exercise
  interventions that target specific components of cognitive performance
  and mental health by leveraging low-cost fitness tracking devices.
\end{abstract}

\section*{Introduction}
Engaging in physical activity (exercise) can improve physical fitness
by increasing muscle strength \citep{RogeEvan93, Lind79, CranEtal13,
  Knut07}, bone density~\citep{ChilEtal12, BassRams94, LaynNels99},
cardiovascular performance~\citep{MaioEtal00, PollEtal00}, lung
capacity~\citep[][although see \cite{RomaEtal16}]{LazoEtal16}, and
endurance~\citep{WilmKnut03}.  Exercise can also improve mental
health~\citep{Ragl90, MikkEtal17, TaylEtal85, DeslEtal09, Call04,
  PaluSchw00, BassSuzu17} and cognitive
performance~\citep{ChanEtal12b, BrisEtal02, EtniEtal06, BassSuzu17}.

The physical benefits of exercise can be explained by stress-responses
of the affected body tissues. For example, skeletal muscles that are
taxed during exercise exhibit stress responses~\citep{MortEtal09} that
can in turn affect their growth or atrophy~\citep{SchiEtal13}.  By
comparison, the benefits of exercise on mental health are less direct.
For example, one hypothesis is that exercise leads to specific
physiological changes, such as increased aminergic synaptic
transmission and endorphin release, which in turn act on
neurotransmitters in the brain~\citep{PaluSchw00}.  Speculatively, if
different exercise regimens lead to different neurophysiological
responses, one might be able to map out a spectrum of signalling and
transduction pathways that are impacted by a given type, duration, and
intensity of exercise in each brain region.  For example, prior work
has shown that exercise increases acetylcholine levels, starting in
the vicinity of the exercised muscles~\citep{ShoeEtal97}.
Acetylcholine is thought to play an important role in memory
formation~\citep[e.g., by modulating specific synaptic inputs from
entorhinal cortex to the hippocampus, albeit in
rodents;][]{PalaEtal21}.  Given the central role that these medial
temporal lobe structures play in memory, changes in acetylcholine
might lead to specific changes in memory formation and retrieval.

In the present study, we hypothesize that (a) different exercise
regimens will have different, quantifiable impacts on cognitive
performance and mental health, and that (b) these impacts will be
consistant across individuals.  To this end, we collected a year of
real-world fitness tracking data from each of 113 participants.  We
then asked each participant to fill out a brief survey in which they
self-evaluated and self-reported several aspects of their mental
health.  Finally, we ran each participant through a battery of memory
tasks, which we used to evaluate their memory performance along
several dimensions.  We searched the data for potential associations
between memory, mental health, and exercise.

@@

\section*{Author contributions}
Concept: J.R.M. and G.M.N.  Experiment implementation and data collection: G.M.N.
Analyses: J.R.M., G.M.N., E.C., and P.C.F.  Writing: J.R.M. with input
from all authors.
